---
title: \pkg{nanotime}&#58; A datetime library with nanosecond precision

# Use letters for affiliations
author:
  - name: Dirk Eddelbuettel
    affiliation: 1
  - name: Leonardo Silvestri
    affiliation: 2
address:
  - code: 1
    address: Department of Statistics, University of Illinois, Urbana-Champaign, IL, USA
  - code: 2
    address: Baltimore, MD, USA

# For footer text  TODO(fold into template, allow free form two-authors)
lead_author_surname: Eddelbuettel, Silvestri

# Place DOI URL or CRAN Package URL here
doi: "https://cran.r-project.org/package=nanotime"

# Abstract
abstract: |
  The \pkg{nanotime} package \citep{CRAN:nanotime} provides a coherent set of
  temporal types and functions with nanosecond precision. The following types
  are provided: point in time, interval (which may have an open or closed
  start/end), period (a human representation of time, such as day, month,
  etc.), and duration. Of particular interest are set and arithmetic
  operations defined on these types, as well as the fact that all
  functionality is designed to correctly handle instances across different
  time zones. The new temporal types are based on R built-in types with an
  efficient implementation, and are suitable for use in `data.frame` and
  `data.table` objects. `nanotime` is also a better choice than the native
  `POSIXct` in most cases where fractional seconds are needed as it provides
  both a finer resolution and additional operations. _Note that this vignette
  should be considered in progress and is not yet finalized. It should, 
  however, provide some useful information and documentation about the package._



# Font size of the document, values of 9pt (default), 10pt, 11pt and 12pt
fontsize: 10pt

# Optional: Force one-column layout, default is two-column
one_column: true

# Optional: Enables lineno mode, but only if one_column mode is also true
#lineno: true

# Optional: Enable one-sided layout, default is two-sided
#one_sided: true

# Optional: Enable section numbering, default is unnumbered
#numbersections: true

# Optional: Specify the depth of section number, default is 5
#secnumdepth: 5

# Optional: Skip inserting final break between acknowledgements, default is false
skip_final_break: true

# Optional: Bibliography
bibliography: nanotime-intro

# Optional: Enable a 'Draft' watermark on the document
watermark: false

# Customize footer, eg by referencing the vignette
footer_contents: "nanotime Vignette"

# Produce a pinp document
output:
  pinp::pinp:
    collapse: true
    keep_tex: false

header-includes: >
  \newcommand{\proglang}[1]{\textsf{#1}}
  \newcommand{\pkg}[1]{\textbf{#1}}

# Required: Vignette metadata for inclusion in a package.
vignette: >
  %\VignetteIndexEntry{Introduction to nanotime}
  %\VignetteKeywords{anytime, date, datetime, conversion}
  %\VignettePackage{anytime}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\usepackage[utf8]{inputenc}
---

```{r initialsetup, include=FALSE}
knitr::opts_chunk$set(cache=FALSE)
library(bit64)
library(nanotime)
library(data.table)
options(width=65)                       # to limit wide rows
```

# Implementation of the temporal types

All new types provided by the \pkg{nanotime} package are implemented using one or more signed 64-bit
integers. As R has no native 64-bit integer type, \pkg{nanotime} relies on the \code{integer64} type
provided by the \pkg{bit64} package \citep{CRAN:bit64}. Using an integer-based representation avoid
the floating point issues associated with the `POSIXct` type using in R, and allows for exact
representation of nanosecond-resolution datetimes in the approximative range of years 1386 to 2554,
which is large enough for most applications. It also allows for exact arithmetic and comparison.

# Time zones

Time zones are only needed in two cases. The first is representation as of course the same time is
represented differently in different time zones across the globe. The second case is for operations
based on calendar time as a year or a day may have different length depending on its location in
time and depending in which time zone it is considered. For example in a time zone where daylight
saving time is observed, there are days of 23 hours and days of 25 hours. Similarly, at historical
time zone offset changes, a year can have a different duration compared to years where no offset
change occurred.


To avoid all ambiguity, a time zone is never associated as an attribute with any of the temporal
types defined in the \pkg{nanotime} package. Any operation that requires a time zone in order to
be sensible has to explicitly specificy the time zone.

# Temporal types

The \pkg{nanotime} package implements four temporal types: time (`nanotime`), interval (`nanoival`),
duration (`nanoduration`) and period (`nanoperiod`). This set of types seems to have become the
consensus for various implementations. It is the set chosen by _Joda Time_ \citep{Joda2019}, the
\pkg{lubridate} package \citep{JSS:lubridate,CRAN:lubridate} (without nanosecond resolution), and a
similar set (minus the interval type) is chosen by the latest Java Date and Time implementation
\citep{Java2014}.

## nanotime

A `nanotime` object represents a point in time with nanosecond resolution. It is implemented as an
S4 class containing the type `integer64` type from the \pkg{bit64} package \citep{CRAN:bit64} to
represents the point in time as the offset in nanoseconds since the "epoch" of 1970-01-01 UTC. A
`nanotime` object not have an associated time zone. It can be displayed in any desired time zone
with the `format` or `print` functions---which both accept a time zone argument to format the point
in time in a human-readable format. Finally, it uses the POSIX definition of time, usually referred
to as _POSIX time_ or _UNIX time_ and defined by \citet{POSIX2016}. This representation is suitable
for many purposes.

A `nanotime` object can be constructed either from an `integer64` or from a `character`:

```{r}
as.nanotime( as.integer64("1580274000000000000") )
as.nanotime( "2020-01-29 13:12:00.000000001 America/New_York" )
as.nanotime( "2020-01-29 13:12:00.000000001-05:00" )
```

For added convenience, short forms are also supported:

```{r}
as.nanotime("2020-01-29 Europe/London")
as.nanotime("2020-01-29 12:01:01 Africa/Lagos")
as.nanotime("2020-01-29 12:01:01.001 America/Tegucigalpa")
```


## nanoival

A `nanoival` object describes an interval in time defined by two `nanotime` objects. Here the first
one defines the interval start time, and the second one the end time. Additionally, two `logical`
variables determine if the interval start and end are open (`TRUE`) or closed (`FALSE`), _i.e._ if
the start and end of the interval are excluded (`TRUE`) or included (`FALSE`) in the interval.

A `nanoival` object can either be constructed with two `nanotime` and two `logical` objects, or it
can be constructed from a `character` string. The string format uses the '-' and '+' signs at the
beginning and end to indicate if the interval start and end are, respectively, open or closed. By
default, an interval has a closed start and an open end.

```{r}
as.nanoival("-2009-01-01 13:12:00 America/New_York -> 2009-02-01 15:11:03 America/New_York+")

start <- nanotime("2009-01-01 13:12:00 America/New_York")
end   <- nanotime("2009-02-01 15:11:00 America/New_York")

nanoival(start, end)                   # by default sopen=F,eopen=T
nanoival(start, end, sopen=FALSE, eopen=TRUE)
```

Accessors for all four components of a `nanoival` objects are provided:

```{r, nanoival_accessors}
ival  <- as.nanoival("-2009-01-01 UTC -> 2009-02-01 UTC+")
nanoival.start(ival)
nanoival.end(ival)
nanoival.sopen(ival)
nanoival.eopen(ival)
```

## nanoduration

A `nanoduration` object is represented by a simple count of nanoseconds, which may be also be
negative.

```{r}
nanoduration(hours=1, minutes=1, seconds=1, nanoseconds=1)
as.nanoduration("00:00:01")
as.nanoduration("-00:00:01")
as.nanoduration("100:00:00")
as.nanoduration("00:00:00.000_000_001")
```

## nanoperiod

A `nanoperiod` object represents the calendar or "business" view of a duration with the concepts of
month and day. The exact duration of a period is unknown until it is anchored to a point in time and
associated with a time zone. This is due to two reasons: first because a month has variable length,
and secondly because it might span a daylight saving time change.

Each `nanoperiod` object is composed of two parts: a months and days part, and a duration. Note that
these components may have opposite signs.

For convenience, the constructor syntax allows specifying years and weeks, but they are converted to
their representation in months and days.

```{r}
as.nanoperiod("1y1m1w1d/01:01:01.000_000_001")
nanoperiod(months=13, days=-1, duration="01:00:00")
```

Accessors for `nanoperiod` components are provided:

```{r}
ones <- as.nanoperiod("1y1m1w1d/01:01:01.000_000_001")
nanoperiod.month(ones)
nanoperiod.day(ones)
nanoperiod.nanoduration(ones)
```

# Set operations

The set operations `intersect`, `union` and `setdiff` are provided and offer a more rigorous
handling of temporal interval types. Additionally, `intersect` and `setdiff` have counterpart
functions `intersect.idx` and `setdiff.idx` that, instead of computing a new set, return the index
of the set. Finally, the operator `%in%` is overloaded so as to provide a convenient intersection
shorthand particularly suitable for the subsetting of time-series stored as `data.table` columns.

Here are some examples:

```{r, setops}
ni1 <- c(as.nanoival("+2013-01-01+00:00          -> 2014-01-01+00:00-"),
         as.nanoival("+2015-01-01T12:00:01+00:00 -> 2016-01-01+00:00-"),
         as.nanoival("+2017-01-01+00:00          -> 2018-01-01+00:00-"))
ni2 <-   as.nanoival("-2013-02-02+00:00          -> 2015-06-10+00:00+")
intersect(ni1, ni2)
union(ni1, ni2)         
setdiff(ni1, ni2)
```


# Functions and operations

## Arithmetic

The standard expected arithmetic operations are defined for various types if and when these
operation are sensible. In particular one can add (or subtract) a `period` or a `duration` to and
from a `nanotime` or a `nanoival`. One can also multiply and divide `period` and `duration` objects
by a scalar.

```{r, arith}
as.nanotime("2020-03-07 01:03:28 America/Los_Angeles") + 999
as.nanotime("2020-03-07 12:03:28+00:00") + as.nanoduration("24:00:00")
## daylight saving time transition:
plus(as.nanotime("2020-03-07 12:03:28+00:00"), as.nanoperiod("1d"), "America/Los_Angeles")

as.nanoduration("24:00:00")/3
-as.nanoduration("24:00:00")
```

## Comparison

Compare operations are mostly straightforward except maybe for `nanoival` which is ordered by its
start `nanotime`. If both starts are equal, a closed start comes before an open start. If both
`sopen` are the same, then the comparison happens on the end of the `nanotime`, with a shorter
interval coming before a longer one. `nanoperiod` objects do not have a meaningful ordering and
therefore remain unordered.

```{r, compare}
as.nanoival("+2020-04-03 00:12:00 UTC -> 2020-04-04 00:12:00 UTC-") <
    as.nanoival("-2020-04-03 00:12:00 UTC -> 2020-04-04 00:12:00 UTC-")
nanotime(1) <= nanotime(2)
as.nanoduration(1) > as.nanoduration(2)
```

## Sequence Generation

Sequence generation is provided for `nanotime` and `nanoival` objects. The increment can either be a
`nanoduration` or a `nanoperiod`. Since a period is sensitive to the time zone in which the operation
takes place, the additional `tz` argument must be provided to `seq` when operating on `nanoperiod`
objects as in the second example.

```{r, seq_generation}
seq(nanotime("2020-03-28+00:00"), by=as.nanoduration("24:00:00"), length.out=3)
seq(nanotime("2020-03-28+00:00"), by=as.nanoperiod("1d"), length.out=3, tz="Europe/London")

ival <- as.nanoival("+2020-03-28T13:00:00+00:00 -> 2020-03-28T15:00:00+00:00-")
print(seq(ival, by=as.nanoperiod("1m"), length.out=3, tz="Europe/London"), tz="Europe/London")

```

Note that `nanoperiod` is correct with respect to time zone calculations, even on the rare hourly
events where a transition occurs from a time zone offset with an hourly difference to a time zone
offset with a half-hourly difference.

``` {r}
print(seq(as.nanotime("2006-04-14 22:00:00 Asia/Colombo"),
          by=as.nanoperiod("01:00:00"),
          length.out=4,
          tz="Asia/Colombo"),
      tz="Asia/Colombo")
```

## Year, Month and Day

Utilities are provided for obtaining in numerical format the day of the week (`nano_wday`), the day
of the month (`nano_mday`), the month (`nano_month`) and the year (`nano_year`) from a given
`nanotime`. Remember that a time zone is never stored with a `nanotime` and therefore, to have
meaning, all the functions take as second argument the time zone required for this computation. Note
that the convention for the day of the week is a count from 0 to 6, with 0 falling on Sunday.

```{r, year_month_day}
tm  <- as.nanotime("2019-12-31 20:00:00", tz="UTC")
nano_wday(tm, "Australia/Melbourne")
nano_wday(tm, "America/New_York")
nano_mday(tm, "Africa/Nairobi")
nano_month(tm, "Indian/Reunion")
nano_year(tm, "Asia/Irkutsk")
```

## Rounding Operations

The functions `nano_floor` and `nano_ceiling` are provided in order to perform rounding to an
arbitrary precision. An `origin` argument of type `nanotime` can be optionally specified to offer
full control over the reference chosen for the rounding.

These functions are also to be understood in the context of vectors of `nanotime` objects where the
precision defines a grid interval. These functions will pick a reasonable reference for the
alignment. In particular, when using a `nanoperiod`, the functions will check if a precision is a
multiple of a larger unit. If so, the rounding will happen with the larger unit as origin. For
instance, if the precision is 6 hours---a multiple of a day---the rounding will be performed in such
a way as to align the vector within a day, _i.e._ the rounding will be done at hours 0, 6, 12
and 18. On the other hand, if the origin is explicitly specified, then it is this value that will be
taken as starting point for the rounding. For instance, if the origin is set to `2020-04-27
23:57:04` then the rounding will be done at 23:57:04, 05:57:04, 11:57:04 and 17:57:04.

```{r, rounding ops}
nano_floor(as.nanotime("2020-04-27 23:57:04.123456678 UTC"),
           as.nanoduration("00:00:00.001"))
nano_ceiling(as.nanotime("2020-04-27 23:57:04.123456678 UTC"),
             as.nanoduration("00:00:00.001"))

nano_floor(as.nanotime("2020-04-27 23:57:04 UTC"), as.nanoperiod("06:00:00"), tz="UTC")
nano_ceiling(as.nanotime("2020-04-27 23:57:04 UTC"), as.nanoperiod("06:00:00"), tz="UTC")

nano_floor(as.nanotime("2020-04-27 23:57:04 America/New_York"),
           as.nanoperiod("1m"), tz="America/New_York")
nano_ceiling(as.nanotime("2020-04-27 23:57:04 America/New_York"),
             as.nanoperiod("1m"), tz="America/New_York")
```

# Use with data.frame and data.table 


All the new types introduced by the \pkg{nanotime} package are compatible with `data.frame` and
`data.table` \citep{CRAN:data.table} objects. By having an ordered `nanotime` column it is thus easy
to define a time-series. One can then use `nanoival` subsetting.

```{r, data.table}

idx <- seq(nanotime("2020-04-02+00:00"), by=as.nanoperiod("1d"), length.out=20, tz="UTC")
dt <- data.table(idx, v1=1:20, v2=c(TRUE, FALSE))
ival <- as.nanoival(c("+2020-04-05 UTC -> 2020-04-07 UTC+",
                      "+2020-04-15 UTC -> 2020-04-17 UTC+"))
dt[idx %in% ival]

```

Use of the rounding functions makes it possible to perform aggregations on `data.table` instances
utilising its powerful 'group-by' operator `by`:

```{r, aggregations}
idx <- seq(as.nanotime("2020-03-08 UTC"), as.nanotime("2020-03-10 UTC"),
           by=as.nanoduration("00:01:00"))
dt <-  data.table(idx, a=1:length(idx))
dt[, .(mean=mean(a)), by=nano_ceiling(idx, as.nanoduration("06:00:00"))]
```


# Input and Output Format

## nanotime

The input and output format default to "%Y-%m-%dT%H:%M:%EXS%Ez" where the 'X' specifies a
variable number of digits for the nanosecond portion. When no overriding format is defined, the
output will include only the relevant nanotime precision for the vector without right-padded zeros
as shown in the following example:

```{r, ioex1}
format(as.nanotime("2020-12-12T00:00:00.000000000+00:00"))
format(as.nanotime("2020-12-12T00:00:00.123000000+00:00"))
format(as.nanotime("2020-12-12T00:00:00.123456000+00:00"))
format(as.nanotime("2020-12-12T00:00:00.123456789+00:00"))
```

Details of the format specification are provided by the documentation for the underlying CCTZ
library by \citet{github:cctz} which is deployed here via the \pkg{RcppCCTZ} package
\citep{CRAN:RcppCCTZ}. When no overriding format is defined, the parsing has some flexibility and
the time portion can be omitted. Additionally, the separator `'_'` can be used to separate
nanosecond groups of 3. So the following examples will parse correctly:

```{r, ioex2}
as.nanotime("2020-04-03 UTC")
as.nanotime("2020-04-03T12:23:00 UTC")
as.nanotime("2020-04-03T12:23:00.1 UTC")
as.nanotime("2020-04-03T12:23:00.123 UTC")
as.nanotime("2020-04-03T12:23:00.123356789 UTC")
as.nanotime("2020-04-03T12:23:00.123_356_789 UTC")

```

Date separators can be `' '`, `'-'` and `'/'` whereas the separator between date and time can be
`'T`' or `' '`:

```{r, ioex3}
as.nanotime("2020 04 03 UTC")
as.nanotime("2020/04/03 UTC")
as.nanotime("2020-04-03T12:23:00 UTC")
```

## nanoival

The output format for `nanoival` objects is based on `nanotime` as a `nanoival` object is composed
of both a `nanotime` start and end object as well as two booleans that indicate if the boundaries of
the interval are open or closed. This open and closed is indicated by prefixing and postfixing with
the the characters `'-'` and `'+'`. The start and end are separated by `'->'`. Two examples follow.

```{r, ioex4}
as.nanoival("+2020-12-12 UTC -> 2020-12-13 UTC-")
as.nanoival("-2020-12-12T00:00:01.123 America/New_York -> 2020-12-14+00:00+")
```

## nanoduration

The output format for a `nanoduration` object is immutable, and identical to the
hour/minute/second/nanosecond portion of a `nanotime` object:

```{r, ioex5}
as.nanoduration("12:23:00")
as.nanoduration("12:23:00.1")
as.nanoduration("12:23:00.123")
as.nanoduration("12:23:00.123356789")
as.nanoduration("12:23:00.123_356_789")
```

## nanoperiod

The output format for `nanoperiod` is also immutable. It is composed of two parts. First comes a
month and day part, which is followed by a `nanoduration` part that is separated by `'/'`. In input,
years, months, weeks and days are specified with a signed integer following, respectively by the
letters `'y'`,`'m'`,`'w'`,`'d'`. The `nanoduration` that composes the second part is specified like
for a standalone `nanoduration`. Each of these two parts is optional. In output, only months and
days are specified as years can be expressed as 12 months and weeks as 7 days. Here are some
examples:

```{r, ioex6}
as.nanoperiod("1y1m1w1d/00:00:00.123") 
as.nanoperiod("-2y")
as.nanoperiod("00:00:00.123")
```

# Technical Details

All new types provided in this package are built with S4 classes containing an R primitive type that
is then reinterpreted. `nanotime` and `duration` are (indirectly) based on `double` via the type
`integer64` from the \pkg{bit64} package written by \citet{CRAN:bit64}, whereas `nanoival` and
`period` are based on `complex`, which allows the storage of 128 bits.  Time-zone conversion and
calculation as well general time operations rely on the \pkg{CCTZ} library \citep{github:cctz}
interfaced by \pkg{RcppCCTZ} \citep{CRAN:RcppCCTZ}. Interfacing to and from C++ is done using the
\pkg{Rcpp} package \citep{JSS:Rcpp,CRAN:Rcpp}.


# Performance

The `as.POSIXct` function in R provides a useful baseline as it is also implemented in compiled
code. The `fastPOSIXct` function from the \pkg{fasttime} package \citep{CRAN:fasttime} excels at
converting one (and only one) input format _fast_ to a (UTC-only) datetime object. A simple
benchmark converting 100 input strings 10,000 times shows that the `nanotime` constructor is much
closer to the performance of the optimal `fastPOSIXct` parser than to the more general `as.POSIXct`
converter, see Table \ref{tab:speed} for details.


```{r knitr_speed, echo=FALSE}
df <- read.table(stringsAsFactors=FALSE, text="
                     test replications elapsed relative
3             as.nanotime        10000   0.304    2.895
5 as.nanotime_with_format        10000   0.377    3.590
4     as.nanotime_with_tz        10000   0.397    3.781
1              as.POSIXct        10000   1.698   16.171
2             fastPOSIXct        10000   0.105    1.000
")
knitr::kable(df, "latex", booktabs=TRUE, row.names=FALSE,
             caption="\\label{tab:speed}Comparison of as.nanotime, base R and fasttime")
```

# Summary

We describe the \pkg{nanotime} package which offers a coherent set of types and operations with
nanosecond precision.

We show that the \pkg{nanotime} package provides the building blocks to build more complicated and
interesting functions, in particular within the context of `data.table` time-series.

# Appendix

The benchmark results shown in table \ref{tab:speed} are based on the code included below, and
obtained via execution under R version 4.0.2 running under Ubuntu 20.04 with Linux kernel 5.4.0-39
on an Intel i7-8700K CPU.

``` {r eval=FALSE}
library(nanotime)
library(rbenchmark)
library(fasttime)


x_posixct       <- rep("2020-03-19 22:55:23", 100)
x_nanotime      <- rep("2020-03-19 22:55:23.000000001+00:00", 100)
x_nanotime_tz   <- rep("2020-03-19 22:55:23.000000001 America/New_York", 100)
x_nanotime_cctz <- rep("03-19-2020 22:55:23.000000001+00:00", 100)

benchmark(
    "as.POSIXct" = { x <- as.POSIXct(x_posixct) },
    "fastPOSIXct" = { x <- fastPOSIXct(x_posixct) },
    "as.nanotime" = { x <- as.nanotime(x_nanotime) },
    "as.nanotime with tz" = { x <- as.nanotime(x_nanotime_tz) },
    "as.nanotime with format" = { x <- as.nanotime(x_nanotime_tz, format="%m-%d-%YT%H:%M:%E9S%Ez") },
    replications = 10000,
    columns = c("test", "replications", "elapsed", "relative"))
```
